---
title: "Model comparisons"
output:
  html_document:
    theme: sandstone
    highlight: tango
    toc: true
    toc_float: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Loading packages
```{r}
library(tidyverse)
library(rptR)
library(gridExtra)
```

# Loading data

```{r}
rf_avg_predictions <- read_csv("../1_data/rf_avg_predictions.csv")
enr_avg_predictions <- read_csv("../1_data/enr_avg_predictions.csv")
gpm_avg_predictions <- read_csv("../1_data/gpm_avg_predictions.csv")
```

# Random forest
```{r, fig.height=4, fig.width=6}
# Plot
rf_plot <- ggplot(rf_avg_predictions, aes(x = age_year, y = avg_bio_age)) +
  geom_point(alpha = 0.5) +
  stat_smooth(color = "firebrick", method = "lm", se=F, size=2, alpha=0.5) +
  theme_bw() +
  geom_abline(intercept = 0, slope = 1, lty = 2) +
  scale_x_continuous(breaks = seq(2, 28, by = 2)) +
  scale_y_continuous(limits = c(4, 18), breaks = seq(2, 28, by = 2)) +
  xlab("Chronological age (years)") +
  ylab("Predicted age (years)") +
  theme(axis.text=element_text(size=10),
        axis.title=element_text(size=12))

# determining the R2
r_rf <- cor(rf_avg_predictions$age_year, rf_avg_predictions$avg_bio_age, method = c("pearson"))
r_rf
r_rf^2

# median error
# median difference between bio age and age
median(abs(rf_avg_predictions$avg_bio_age - rf_avg_predictions$age_year))
mean(abs(rf_avg_predictions$avg_bio_age - rf_avg_predictions$age_year))
range(rf_avg_predictions$avg_bio_age - rf_avg_predictions$age_year)


# slope
rf_mod <- lm(avg_bio_age ~ age_year, data=rf_avg_predictions)
summary(rf_mod)
rf_mod$coefficients
```

## Repeatability

more generally referred to as the intra-class correlation (ICC)
https://cran.r-project.org/web/packages/rptR/vignettes/rptR.html

The repeatability coefficient is a precision measure which represents the value below which the absolute difference between two repeated test results may be expected to lie with a probability of 95%.

INTERPRETATION
https://onlinelibrary.wiley.com/doi/full/10.1002/evl3.40
"Individual repeatability (R), defined as the proportion of observed variance attributable to among-individual differences"

### Residuals
```{r}
# Taking the residuals of rf_mod
rf_avg_predictions$bio_age_resid = residuals(rf_mod)

rf_avg_predictions$bio_age_resid = as.numeric(rf_avg_predictions$bio_age_resid)

# ensuring there is no age association
rf_avg_predictions %>% 
  ggplot(aes(x=age_year, y=bio_age_resid)) +
  geom_point() +
  geom_smooth(method="lm")


# analysis
rf_rpt <- rpt(bio_age_resid ~ (1 | anon_sname), grname = "anon_sname", 
              data = rf_avg_predictions, datatype = "Gaussian", 
    nboot = 0, npermut = 0)
print(rf_rpt)

# bootstrapping to get CI
rf_rpt_bs <- rpt(bio_age_resid ~ (1 | anon_sname), grname = "anon_sname", 
                 data = rf_avg_predictions, datatype = "Gaussian", 
    nboot = 1000, npermut = 0)
print(rf_rpt_bs)
plot(rf_rpt_bs, cex.main = 1)
```

#### Figure 2B
```{r, fig.height=5, fig.width=8}
# finding top 25 sampled individuals
top_sampled_id <- rf_avg_predictions %>% 
  group_by(anon_sname) %>% 
  dplyr::summarize(n = n(),
            median = median(bio_age_resid)) %>% 
  arrange(desc(n)) %>% 
  slice(1:25)

repeatability <- rf_avg_predictions %>% 
  filter(anon_sname %in% top_sampled_id$anon_sname) %>%
  left_join(top_sampled_id, by = "anon_sname") %>% 
  mutate(anon_sname = fct_reorder(as.factor(anon_sname), median)) %>%
  ggplot() +
  geom_jitter(aes(x = anon_sname, y = bio_age_resid, color = anon_sname)) +
  geom_boxplot(aes(x=anon_sname, y = bio_age_resid), alpha = 0.1, outlier.shape = NA) +
  theme_bw() +
  ylab(bquote(''*'Age'[P]*', controlling for ' *'Age'[C]*'')) +
  xlab("Individual identity") +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        legend.position = "none",
        axis.title = element_text(size = 14),
        axis.text = element_text(size = 10))
repeatability

```



### Delta age
```{r}
# delta age 
rf_avg_predictions <- rf_avg_predictions %>% 
  mutate(delta_age = avg_bio_age - age_year)

rf_avg_predictions %>% 
  ggplot() +
  geom_histogram(aes(x=delta_age)) +
  theme_bw()

rf_avg_predictions %>% 
  ggplot(aes(x=age_year, y=delta_age)) +
  geom_point() +
  geom_smooth() +
  theme_bw()


# analysis
rf_rpt_deltaage <- rpt(delta_age ~ age_year + (1 | anon_sname), grname = "anon_sname", 
              data = rf_avg_predictions, datatype = "Gaussian", 
    nboot = 0, npermut = 0)
print(rf_rpt_deltaage)

# bootstrapping to get CI
rf_rpt_bs_deltaage <- rpt(delta_age ~ age_year + (1 | anon_sname), grname = "anon_sname", 
                 data = rf_avg_predictions, datatype = "Gaussian", 
    nboot = 1000, npermut = 0)
print(rf_rpt_bs_deltaage)
plot(rf_rpt_bs_deltaage, cex.main = 1)
```

# ENR
```{r}
# Plot
ggplot(enr_avg_predictions, aes(x = age_year, y = avg_bio_age)) +
  geom_point(alpha = 0.25) +
  stat_smooth(fill = "firebrick", color = NA, method = "lm") +
  geom_abline(intercept = 0, slope = 1, lty = 2) +
  scale_x_continuous(breaks = seq(2, 28, by = 2)) +
  scale_y_continuous(breaks = seq(2, 28, by = 2)) +
  labs(x = "Chronological Age", y = "Biological Age") +
  theme_bw()

# determining the R2
r_enr <- cor(enr_avg_predictions$age_year, enr_avg_predictions$avg_bio_age, method = c("pearson"))
r_enr
r_enr^2

# median error
# median difference between bio age and age
median(abs(enr_avg_predictions$avg_bio_age - enr_avg_predictions$age_year))
mean(abs(enr_avg_predictions$avg_bio_age - enr_avg_predictions$age_year))
range(enr_avg_predictions$avg_bio_age - enr_avg_predictions$age_year)


# slope
enr_mod <- lm(avg_bio_age ~ age_year, data=enr_avg_predictions)
summary(enr_mod)
enr_mod$coefficients
```

## Repeatability

### Residuals
```{r}
# Taking the residuals of enr_mod
enr_avg_predictions$bio_age_resid = residuals(enr_mod)

# ensuring there is no age association
enr_avg_predictions %>% 
  ggplot(aes(x=age_year, y=bio_age_resid)) +
  geom_point() +
  geom_smooth(method="lm")


# analysis
enr_rpt <- rpt(bio_age_resid ~ (1 | anon_sname), grname = "anon_sname", 
              data = enr_avg_predictions, datatype = "Gaussian", 
    nboot = 0, npermut = 0)
print(enr_rpt)

# bootstrapping to get CI
enr_rpt_bs <- rpt(bio_age_resid ~ (1 | anon_sname), grname = "anon_sname", 
                 data = enr_avg_predictions, datatype = "Gaussian", 
    nboot = 1000, npermut = 0)
print(enr_rpt_bs)
plot(enr_rpt_bs, cex.main = 1)
```

### Delta age
```{r}
# delta age 
enr_avg_predictions <- enr_avg_predictions %>% 
  mutate(delta_age = avg_bio_age - age_year)

enr_avg_predictions %>% 
  ggplot() +
  geom_histogram(aes(x=delta_age)) +
  theme_bw()

enr_avg_predictions %>% 
  ggplot(aes(x=age_year, y=delta_age)) +
  geom_point() +
  geom_smooth() +
  theme_bw()


# analysis
enr_rpt_deltaage <- rpt(delta_age ~ age_year + (1 | anon_sname), grname = "anon_sname", 
              data = enr_avg_predictions, datatype = "Gaussian", 
    nboot = 0, npermut = 0)
print(enr_rpt_deltaage)

# bootstrapping to get CI
enr_rpt_bs_deltaage <- rpt(delta_age ~ age_year + (1 | anon_sname), grname = "anon_sname", 
                 data = enr_avg_predictions, datatype = "Gaussian", 
    nboot = 1000, npermut = 0)
print(enr_rpt_bs_deltaage)
plot(enr_rpt_bs_deltaage, cex.main = 1)
```

# GPM
```{r}
# Plot
ggplot(gpm_avg_predictions, aes(x = age_year, y = avg_bio_age)) +
  geom_point(alpha = 0.25) +
  stat_smooth(fill = "firebrick", color = NA, method = "lm") +
  geom_abline(intercept = 0, slope = 1, lty = 2) +
  scale_x_continuous(breaks = seq(2, 28, by = 2)) +
  scale_y_continuous(breaks = seq(2, 28, by = 2)) +
  labs(x = "Chronological Age", y = "Biological Age") +
  theme_bw()

# determining the R2
r_gpm <- cor(gpm_avg_predictions$age_year, gpm_avg_predictions$avg_bio_age, method = c("pearson"))
r_gpm
r_gpm^2

# median error
# median difference between bio age and age
median(abs(gpm_avg_predictions$avg_bio_age - gpm_avg_predictions$age_year))
mean(abs(gpm_avg_predictions$avg_bio_age - gpm_avg_predictions$age_year))
range(gpm_avg_predictions$avg_bio_age - gpm_avg_predictions$age_year)

# slope
gpm_mod <- lm(avg_bio_age ~ age_year, data=gpm_avg_predictions)
summary(gpm_mod)
gpm_mod$coefficients
```

## Repeatability

### Residuals
```{r}
# Taking the residuals of enr_mod
gpm_avg_predictions$bio_age_resid = residuals(gpm_mod)

# ensuring there is no age association
gpm_avg_predictions %>% 
  ggplot(aes(x=age_year, y=bio_age_resid)) +
  geom_point() +
  geom_smooth(method="lm")


# analysis
gpm_rpt <- rpt(bio_age_resid ~ (1 | anon_sname), grname = "anon_sname", 
              data = gpm_avg_predictions, datatype = "Gaussian", 
    nboot = 0, npermut = 0)
print(gpm_rpt)

# bootstrapping to get CI
gpm_rpt_bs <- rpt(bio_age_resid ~ (1 | anon_sname), grname = "anon_sname", 
                 data = gpm_avg_predictions, datatype = "Gaussian", 
    nboot = 1000, npermut = 0)
print(gpm_rpt_bs)
plot(gpm_rpt_bs, cex.main = 1)
```

### Delta age
```{r}
# delta age 
gpm_avg_predictions <- gpm_avg_predictions %>% 
  mutate(delta_age = avg_bio_age - age_year)

gpm_avg_predictions %>% 
  ggplot() +
  geom_histogram(aes(x=delta_age)) +
  theme_bw()

gpm_avg_predictions %>% 
  ggplot(aes(x=age_year, y=delta_age)) +
  geom_point() +
  geom_smooth() +
  theme_bw()


# analysis
gpm_rpt_deltaage <- rpt(delta_age ~ age_year + (1 | anon_sname), grname = "anon_sname", 
              data = gpm_avg_predictions, datatype = "Gaussian", 
    nboot = 0, npermut = 0)
print(gpm_rpt_deltaage)

# bootstrapping to get CI
gpm_rpt_bs_deltaage <- rpt(delta_age ~ age_year + (1 | anon_sname), grname = "anon_sname", 
                 data = gpm_avg_predictions, datatype = "Gaussian", 
    nboot = 1000, npermut = 0)
print(gpm_rpt_bs_deltaage)
plot(gpm_rpt_bs_deltaage, cex.main = 1)
```

# Looking at correlations

## Between bio age
```{r}
# rf and enr
cor(rf_avg_predictions$avg_bio_age, enr_avg_predictions$avg_bio_age)
# rf and gpm
cor(rf_avg_predictions$avg_bio_age, gpm_avg_predictions$avg_bio_age)
# enr and gpm
cor(enr_avg_predictions$avg_bio_age, gpm_avg_predictions$avg_bio_age)


# Plotting them against each other
data <- rf_avg_predictions %>% 
  rename(rf_bio_age = avg_bio_age) %>% 
  left_join(select(enr_avg_predictions, anon_sname, age_year, enr_bio_age = avg_bio_age), by=c("anon_sname", "age_year")) %>% 
  left_join(select(gpm_avg_predictions, anon_sname, age_year, gpm_bio_age = avg_bio_age), by=c("anon_sname", "age_year"))


data %>% 
  ggplot(aes(x=rf_bio_age, y=enr_bio_age)) +
  geom_point() +
  geom_smooth()

data %>% 
  ggplot(aes(x=rf_bio_age, y=gpm_bio_age)) +
  geom_point() +
  geom_smooth()

data %>% 
  ggplot(aes(x=gpm_bio_age, y=enr_bio_age)) +
  geom_point() +
  geom_smooth()

```

## Between delta age
```{r}
# rf and enr
cor(rf_avg_predictions$delta_age, enr_avg_predictions$delta_age)
# rf and gpm
cor(rf_avg_predictions$delta_age, gpm_avg_predictions$delta_age)
# enr and gpm
cor(enr_avg_predictions$delta_age, gpm_avg_predictions$delta_age)


# Plotting them against each other
data2 <- rf_avg_predictions %>% 
  rename(rf_delta_age = delta_age) %>% 
  left_join(select(enr_avg_predictions, anon_sname, age_year, enr_delta_age = delta_age), by=c("anon_sname", "age_year")) %>% 
  left_join(select(gpm_avg_predictions, anon_sname, age_year, gpm_delta_age = delta_age), by=c("anon_sname", "age_year"))


data2 %>% 
  ggplot(aes(x=rf_delta_age, y=enr_delta_age)) +
  geom_point() +
  geom_smooth()

data2 %>% 
  ggplot(aes(x=rf_delta_age, y=gpm_delta_age)) +
  geom_point() +
  geom_smooth()

data2 %>% 
  ggplot(aes(x=gpm_delta_age, y=enr_delta_age)) +
  geom_point() +
  geom_smooth()

```



# Figure S5
```{r, fig.width=10, fig.height=4.5}
rf <- ggplot(rf_avg_predictions, aes(x = age_year, y = avg_bio_age)) +
  geom_point(alpha = 0.5) +
  stat_smooth(color = "firebrick", method = "lm", se=F, size=2, alpha=0.5) +
  theme_bw() +
  geom_abline(intercept = 0, slope = 1, lty = 2) +
  scale_x_continuous(breaks = seq(2, 28, by = 2)) +
  scale_y_continuous(limits = c(4, 20), breaks = seq(2, 28, by = 2)) +
  xlab(bquote('Chronological age ('*'Age'[C]*')')) +
  ylab(bquote('Predicted age ('*'Age'[P]*')')) +
  ggtitle("Random forest regression") +
  theme(axis.text=element_text(size=10),
        axis.title=element_text(size=12)) +
  labs(tag="   ")

enr <- ggplot(enr_avg_predictions, aes(x = age_year, y = avg_bio_age)) +
  geom_point(alpha = 0.5) +
  stat_smooth(color = "firebrick", method = "lm", se=F, size=2, alpha=0.5) +
  theme_bw() +
  geom_abline(intercept = 0, slope = 1, lty = 2) +
  scale_x_continuous(breaks = seq(2, 28, by = 2)) +
  scale_y_continuous(limits = c(4, 20), breaks = seq(2, 28, by = 2)) +
  xlab(bquote('Chronological age ('*'Age'[C]*')')) +
  ylab(bquote('Predicted age ('*'Age'[P]*')')) +
  ggtitle("Elastic net regression") +
  theme(axis.text=element_text(size=10),
        axis.title=element_text(size=12)) +
  labs(tag="   ")

gpm <- ggplot(gpm_avg_predictions, aes(x = age_year, y = avg_bio_age)) +
  geom_point(alpha = 0.5) +
  stat_smooth(color = "firebrick", method = "lm", se=F, size=2, alpha=0.5) +
  theme_bw() +
  geom_abline(intercept = 0, slope = 1, lty = 2) +
  scale_x_continuous(breaks = seq(2, 28, by = 2)) +
  scale_y_continuous(limits = c(4, 20), breaks = seq(2, 28, by = 2)) +
  ggtitle("Gaussian process regression") +
  xlab(bquote('Chronological age ('*'Age'[C]*')')) +
  ylab(bquote('Predicted age ('*'Age'[P]*')')) +
  theme(axis.text=element_text(size=10),
        axis.title=element_text(size=12)) +
  labs(tag="   ")

rf_en_gp <- grid.arrange(rf, enr, gpm, nrow=1)
```

# Figure S7

```{r, fig.width=7, fig.height=6}
# data are from Mauna's paper
other_mod_compare <- data.frame(model = c("Epigenetic clock", "Dentine exposure", "Microbiome clock", "Immature BMI",
                                   "Adult BMI","Flow cytometry", "Blood smear", "NPB clock"),
                         r2 = c(0.60, 0.85, 0.49, 0.33, 0.30, 0.00, 0.04, 0.51),
                         median_error = c(1.62, 1.12, 2.15, 2.30, 3.25, 4.27, 2.00, 2.33))

other_mod_compare <- other_mod_compare %>% 
  mutate(model=fct_relevel(model,c("NPB clock",
                                   "Dentine exposure",
                                   "Epigenetic clock",
                                   "Microbiome clock",
                                   "Immature BMI",
                                   "Adult BMI",
                                   "Blood smear",
                                   "Flow cytometry"))) %>%
 arrange(model)


R2 <- other_mod_compare %>% 
  ggplot(aes(x=model, y=r2)) +
  geom_point(size = 2) +
  theme_bw() +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 10)) +
  xlab("") +
  ylab(bquote(R^2)) +
  theme(axis.text=element_text(size=10),
        axis.title=element_text(size=12))

med_error <- other_mod_compare %>% 
  ggplot(aes(x=model, y=median_error)) +
  geom_point(size = 2) +
  theme_bw() +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 10)) +
  xlab("") +
  ylab("Median error") +
  theme(axis.text=element_text(size=10),
        axis.title=element_text(size=12))

other_mod_comparison_plot <- grid.arrange(R2, med_error, nrow=2)
```


